Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-09-12 manuscript/manuscript.Rmd
finished the 2nd round of revision.

---
title: "Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
abstract: |
  ```{=latex}
  This paper presents the outline of the Perception-Process-Product ("Triple P") conceptual framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes the amplification of multi-level variation and the identification of interacting relationships of variations across the levels of perception, process, and product. Here I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.

  \textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production
  ```
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

Built upon the *Homo economicus* critique and early works in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework. Ontologically speaking, it requires 1) acknowledging the contribution and limitations of actualistic experiments and 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices. These two principles are developed to advocate a pluralistic approach to the explanation of complex variation instead of treating the optimization-based research agenda as a panacea. Epistemologically speaking, the Triple P framework 3) adopts a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.

@@

The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267]. The past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018]. In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.). All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapable of inferring how context-generic principles interact in a particular context as reflected in real-world conditions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be potentially problematic [@liu2023]. In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of cultural transmission. After all, these experimental results can only be as robust as their experimental settings.

On the other hand, actualistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008]. Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals. Interestingly, researchers in cognitive psychology and neuroscience, a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022]. Unlike controlled experiments, variation could be easily observed in actualistic experiments by design. This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with groundstone artifacts as the target products [e.g., @stout2002; @arthur2018]. Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in epdedimiology and economics in recent years [@cunningham2021; @hernan2023; @nichols2007]. Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al. [-@lin2018: 680-681] and the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."

@@

Contemporary practices in experimental archaeology, as manifested by the fact that a majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher [@whittaker1999], tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise". The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989]. When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable [@moore2020: 654]. The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists. However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.

Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61], rather than from the inter-generational transmission of knapping knowledge spanning millions of years. Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers. They tend to come up with ideas that may appear to be counter-intuitive at first glance for academics. One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree[-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology. In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis. It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] famous ethnography. Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers. Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.

Emphasizing variation at its core, the Triple P conceptual framework inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts. The Triple P framework recognizes that experimental archaeology can greatly benefit from diverse perspectives and contributions from multiple stakeholders. By engaging researchers, practitioners, and local communities from different geographical locations, the framework acknowledges the importance of including voices from various cultural backgrounds and contexts [@pargeter2023: 164]. This emphasis on collaboration and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions. Furthermore, the Triple P framework promotes the recognition and value of local knowledge and expertise. It acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage. By involving these local communities in the research process, the framework allows their voices to be heard and their contributions to be acknowledged. This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@douglass2020; @marshall2002].

However, the facilitation of large-scale collaborations faces challenges within the current system of research evaluation. The prevailing practice of attributing credit primarily to the first author and senior (last/corresponding) author in peer-reviewed journal papers hampers the recognition of multiple contributors. This system often overlooks the valuable input of collaborators who may not fit into the traditional authorship structure but have made significant intellectual and practical contributions to the research. To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations.

@@

While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely neglected among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent. One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023]. Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal. This approach is a powerful and elegant yet limited by its normative and analytical orientation, meaning it cannot handle variation very well. To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner. However, this may be biased given that 1) experiments and modeling studies in cognitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practitioners in traditional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].

>>>>

---
title: "Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
abstract: |
  ```{=latex}
  This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes the amplification of multi-level variation and the identification of interacting relationships of variations across the levels of perception, process, and product. Here I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments in the trade-off of causality and generalizability; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.

  \textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; Curse of knowledge; Collaborative knowledge production
  ```
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

Built upon the *Homo economicus* critique and early works in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions. To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework. Ontologically speaking, it requires 1) acknowledging the contribution and limitations of actualistic experiments in the trade-off of causality and generalizability and 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices. These two principles are developed to advocate a pluralistic approach to the explanation of complex variation, which has received more attention from evolutionary anthropology [@antón2017] to cognitive science [@barrett2020], instead of treating the optimization-based research agenda as a panacea. Epistemologically speaking, the Triple P framework 3) adopts a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects. There is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.

@@

The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design across different disciplines [@eren2016; @roe2009: 1266-1267]. In fields known for their development of rigorous and well-controlled experimental methods such as cognitive psychology and neuroscience, researchers have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiments [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022]. In contrast, the past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018]. In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.). All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design can be insufficient in inferring how context-generic principles interact in a particular context as reflected in real-world conditions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies [@liu2023]. In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in cultural transmission and skill development [@proffitt2022]. After all, these experimental results can only be as robust as their experimental settings.

On the other hand, actualistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008]. Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals. Unlike controlled experiments, variation could be easily observed in actualistic experiments by design. This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components do not have analogues in contemporary non-industrial societies [e.g., @stout2002; @arthur2018]. Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in epidemiology and economics in recent years [@cunningham2021; @hernan2023; @nichols2007]. Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al. [-@lin2018: 680-681] and the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."

@@

Contemporary practices in experimental archaeology, as manifested by the fact that a majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher [@whittaker1999], tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise". This psychological term originally refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989], but it has further implications for the sample representativeness in experimental archaeology. When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable [@moore2020: 654]. The existence of this cognitive bias is not inherently bad, and these many years of experience should be appreciated and celebrated by experimental archaeologists. However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.

Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61]. Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers. They tend to come up with ideas that may appear to be counter-intuitive at first glance for academics. One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree [-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology. In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis. It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] famous ethnography. Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers. Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.

Emphasizing variation at its core, the Triple P conceptual framework recognizes that experimental archaeology can greatly benefit from diverse perspectives [@pargeter2023: 164] and thereby inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts. Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage. This emphasis on team efforts and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions. Through ethical collaborations with those knapping practitioners in non-industrial societies in the research process, the framework allows their voices to be heard and their contributions to be acknowledged. This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@douglass2020; @marshall2002].

@@

While systematic behavioral coding methods that are widely used in the study of non-human animal behavior [@fragaszy2018] still largely neglected among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent. One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @haidle2014; @haidle2023]. Cognigram is a graphical representation of the reconstructed behavior behind archaeological artifacts in chronological order of appearance [@haidle2014], which essentially represents an abstracting process of a series of action sequences achieving a similar goal. This approach is a powerful and elegant yet limited by its normative and analytical orientation, meaning it cannot handle variation very well. To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner. However, this may be biased given that 1) experiments and modeling studies in cognitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practitioners in traditional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].

### 2023-09-12 manuscript/manuscript_word.Rmd
finished the 2nd round of revision.

---
title: "Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
abstract: "Abstract: This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes the amplification of multi-level variation and the identification of interacting relationships of variations across the levels of perception, process, and product. Here I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.  \\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production"
date: "`r Sys.Date()`"
output:
  word_document
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
fontsize: 11pt
always_allow_html: true
---

>>>>

---
title: "Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
abstract: "Abstract: This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes the amplification of multi-level variation and the identification of interacting relationships of variations across the levels of perception, process, and product. Here I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments in the trade-off of causality and generalizability; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.  \\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographical analysis; Curse of knowledge; Collaborative knowledge production"
date: "`r Sys.Date()`"
output:
  word_document
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
fontsize: 11pt
always_allow_html: true
---
